---
title: "Lesson 1"
author: "Nicholas Clark"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE,message=FALSE,results="hide",fig.show="hide")
```

# Contact Information

COL Nick Clark

nicholas.clark@westpoint.edu

Thayer Hall Room 229

# Course Overview

\textit{Applied } Statistics

-Course Text/Videos

-Hands on/interactive

-R/Rstudio

-Course Overview

-Graded Events

# Introductions

# DAW, Section Marcher Role


# Berkely Admissions Data

In the early 1970s, the University of California at Berkeley was concerned with possible
discrimination against women in its graduate admissions process. Data about the applicants for
the 1972–73 school year were recorded from several programs, including their sex and whether
or not they were accepted (Bickel & O’Connell, Science, 1975).

```{r}
#install.packages("tidyverse")
library(tidyverse)
berkley.data<-read.table("http://www.isi-stats.com/isi2/data/Berkeley.txt",header=T)
```

Observational Unit:

\vspace{1in}

Response Variable:

\vspace{.5in}

Values Response Variable can take on:

\vspace{.5in}

Visualize

```{r}
ggplot(data=berkley.data,aes(Accepted))+geom_histogram(stat="count")
```

Explore
```{r,include=F}
no.program<-berkley.data %>% select(-Program)
table(no.program)
```
Visualize the table:

```{r}
#install.packages
library(ggmosaic)
ggplot(data=no.program)+
  geom_mosaic(aes(x=product(Accepted,Sex),fill=Accepted))
```
What is the explanatory variable? Are we convinced?
\vspace{.5in}

What would the mosaic plot look like if sex was not associated with acceptance?

\vspace{.5in}

What other explanatory variables might help us predict whether or not someone would be accepted?

\vspace{.4in}

```{r}
ggplot(data=berkley.data)+
  geom_mosaic(aes(x=product(Accepted,Sex),fill=Accepted))+
  facet_grid(Program~.)
```

What are your thoughts looking at this plot?

\vspace{.4in}

How can it be that females have higher acceptance rates than males in Program A AND in Program F but when we combine the two, the overall acceptance rate is noticeably smaller for females?

\vspace{.4in}

In examining gender and acceptance what is \textbf{confounding?}

\vspace{.2in}

What else could contribute to an admissions decision other than sex and program?

\vspace{.5in}

Sources of Variation Diagram:

\vspace{3.in}


Six Steps of a statistical investigation

\begin{itemize}
\item Ask a Research Question

\item Design a study and collect data

\item Explore the data

\item Draw inferences beyond the data

\item Formulate conclusions

\item Look back and ahead

\end{itemize}


Did we do this with the Berkley Data?  What would you tell the college administrators?

\vspace{3.in}


Let's brainstorm how to write a statistical model for the Berkley data



\vspace{3.in}

How would we know if our model is doing a 'good' job?
