---
title: "Lsn 15"
author: "Clark"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Admin

Let's reconsider the Salary Discrimination dataset


```{r}
salary.dat<-read.table("http://www.isi-stats.com/isi2/data/Wages.txt",header=T)
```

Instead of looking at College educated vs not college educated, we now consider the full dataset.

```{r}
levels(salary.dat$educ)
gr.means=salary.dat%>%group_by(educ,race)%>%summarize(mean.salary=mean(wage))

```

What do we see?

```{r}
gr.means$educ<-factor(gr.means$educ,levels=c("belowHS","HS","college","beyondCollege"))
gr.means %>% ggplot(aes(x=educ,y=mean.salary,color=race))+
  geom_line(aes(group=race),lwd=2)+geom_point()
```

\newpage
A statistical model:

\vspace{2.in}

ANOVA table:

\vspace{1.5in}

To fit the model we use:

```{r}
contrasts(salary.dat$race)=contr.sum
contrasts(salary.dat$educ)=contr.sum
inter.lm<-lm(wage~race*educ,data=salary.dat)
coef(inter.lm)
```
Getting the fits is a bit of a pain but we can do it:

\vspace{2.in}

To fit the ANOVA model we note that we are now interested in Type III Sums of squares.


```{r}
library(car)
Anova(inter.lm,type=3)
```

What does this mean?