---
title: "Lsn 8"
author: "Clark"
date: "September 16, 2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Admin

## Power

Recall, in MA206 we tested Hypothesis such as:

$H_0: \mu = \mu_0$ vs $H_a: \mu \neq \mu_0$

Assuming that $H_0$ is true, we created a standardized statistic:

$$ t= \frac{\bar{X}-\mu_0}{s/\sqrt{n}}$$

Which looks like:

\vspace{1.5in}

Using $\alpha=0.05$ our \textit{Rejection Region} can be generated, which gives the values of $t$ that we would conclude $\mu \neq \mu_0$ for.  Note that if $\mu = \mu_a$ we \textit{should} reject $H_a$, however there is a chance that we will not:

\vspace{1.5in}

This region gives the probability of committing a Type-II error for $\mu=\mu_a$.  Statistical power is 1-P(Type-II error) or the probability that the researchers find evidence for the alternative hypothesis when the alternative hypothesis is true.

Note that in order to actually find Power or Type-II error we need to specify a value of $\mu_a$.  Going back to our picture we see:

\vspace{1.5in}

So the value chosen for $\mu_a$ impacts Power, another thing that impacts Power is $n$.

\vspace{1.5in}

For a two-sample or one-sample t-test, finding power in R is straight forward with the command `power.t.test`.  Here we have to input $n$, $\Delta=|\mu_0-\mu_a|$ which can be thought of as how big of an effect do we want to observe, $s$, and/or power.  Note that one of these must be blank and is solved for.  For instance, if we want to observe an effect of 1 and we assume $s=1$, and we have $n=10$ observations for group and we are testing $H_0: \mu_1 = \mu_2$ vs $H_a: \mu_1 \neq \mu_2$ we can run
```{r}

power.t.test(n=10,delta=1,sd=1,power = NULL,type="two.sample",
             alternative="two.sided")
```

And we see that the power associated with this test is 0.56, or the probability of committing a Type-II error in this instance is 0.44.  If we wwant to determine how big of a sample we would need to have 80 \% power we would run:

```{r}

power.t.test(n=NULL,delta=1,sd=1,power=0.8,type="two.sample",
             alternative="two.sided")
```
To see that we would need 17 people in each group.  

The same thing can be done with ANOVA tests using `library(pwr)` which must be installed.

The function `pwr.anova.test` can then be used in a similar way.  The only tricky thing is that it relies on an effect size not given by $R^2$ but rather given by $f=\sqrt{R^2/(1-R^2)}$.  So for instance we have:

```{r,warning=FALSE}
library(pwr)
R2=.1
f=sqrt(R2/(1-R2))
pwr.anova.test(f=f,k=5,n=10)
```

Which matches Table 1.6.4.  We can then change:

```{r,warning=FALSE}
R2=.3
f=sqrt(R2/(1-R2))
pwr.anova.test(f=f,k=5,n=10)
```
Or change People within group
```{r,warning=FALSE}
R2=.1
f=sqrt(R2/(1-R2))
pwr.anova.test(f=f,k=5,n=20)
```
