---
title: "Lsn 24_AY_23"
author: "Clark"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,fig.width=12, fig.height=6)
library(tidyverse)
```

## Admin

Recall we previously looked a pisatcio bleaching where we had three values of air velocity and three values of drying temperature.  Recall that we fit the model:


\vspace{1.in}

Why might we be less than satisfied by this model or this analysis?

\vspace{1.in}

As it turns out, there were three values for air velocity and three values of drying temperature. 

Let's think about designing this study a bit.  Why might our researches have chosen 60, 75, and 90, as our temperatures and 1.5,2, and 2.5 as our air velocity values?

\vspace{.5in}

If our researchers had 45 pistachios to study, how should they allocate pistachios to temperature and air velocity? Why?

\vspace{1.in}


```{r}
pistachio.raw<-read.table("http://www.isi-stats.com/isi2/data/pistachioStudy.txt",header=T)
pistachio.dat <- pistachio.raw %>% mutate(Peroxide=Peroxide..remaining.)%>%
  select(-Peroxide..remaining.)
pistachio.dat<-pistachio.dat %>% select(Temperature,AirVelocity,Peroxide)
```

One way to explore the data is to look at pairs plots

```{r,warning=FALSE, out.width='60%',message=FALSE,fig.align="center"}
library(GGally)
pistachio.dat %>% ggpairs()
```

What is this output showing us?

\vspace{1.in}

```{r,warning=FALSE}
library(rgl)
x=pistachio.dat$Temperature
y=pistachio.dat$AirVelocity
z=pistachio.dat$Peroxide
plot3d(x,y,z)
```

Note that this might lead to the model:


\vspace{1.in}

To fit this, we can fit:

```{r}
pistachio.lm<-lm(Peroxide~AirVelocity+Temperature,data=pistachio.dat)
summary(pistachio.lm)
```

So our fitted model is:

\vspace{1.in}

Our conclusion based on the p values is:

\vspace{1.in}

Let's take a look at the univariate models:

```{r}
uni1.lm<-lm(Peroxide~Temperature,data=pistachio.dat)
summary(uni1.lm)
```

What do we note about the value of temperature?  Is this surprising?

\vspace{.5in}

What about the SE?  Why might this be?

\vspace{.5in}

```{r}
uni2.lm<-lm(Peroxide~AirVelocity,data=pistachio.dat)
summary(uni2.lm)
```

What is the relationship between $R^2$ in the full model and $R^2$ in our two univariate models?  Why might this be?

If we look at $R^2$ of the two univariate models, which covariate explains more variation, Temperature or Air Velocity?

\vspace{.5in}

Looking at $\hat{\beta}_1$ and $\hat{\beta}_2$ which value is bigger?  Is this surprising in light of what we found when looking at $R^2$ values?


\vspace{.5in}

The biggest issue here is we are on different scales.  In order to get rid of scaling we standardize, which is:

\vspace{.5in}

When we have standardized coefficients we can compare values to determine which one has a larger effect.

```{r}
pistachio.std<-pistachio.raw %>%mutate(Peroxide=Peroxide..remaining.)%>%
  select(std.temp,std.air,std.peroxide,Peroxide)
std.lm<-lm(Peroxide~std.temp+std.air,data=pistachio.std)
summary(std.lm)
```

Note here our intercept actually means something.  

\vspace{.5in}


To see if this is a good model, we can examine the residuals:

```{r,out.width='60%',fig.align="center"}
std.lm %>%ggplot(aes(x=.fitted,y=.resid,color=as.factor(std.temp)))+geom_point()
```

Do we have any concerns?  What happens when temp is high?  

\vspace{.5in}

A model with an interaction term:

\vspace{1.in}

For a fixed $X_2$ what happens as we change $X_1$ by 1 unit?  

\vspace{.5in}

Conversely, for a fixed $X_1$ what happens when we change $X_2$ by 1 unit?

\vspace{.5in}

```{r}
inter.lm<-lm(Peroxide~std.temp*std.air,data=pistachio.std)
summary(inter.lm)
```

What's going on here?  What are we testing?  

\vspace{.5in}

Here we see that $R^2$ is still additive

Note that our $\textbf{X}$ matrix becomes:

\vspace{1.in}

Which is nested with the $\textbf{X}$ matrix without an interaction, so we can test whether a model with an interaction is preferable through:

```{r}
anova(std.lm,inter.lm)
```

